---
title: "Analysis of pinger results"
author: "Ben Anderson (@dataknut)"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    theme: journal
    toc: yes
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
# set default echo to FALSE (code not in output)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig_caption = TRUE)
knitr::opts_chunk$set(tidy = TRUE)
```

# About this document

Last run: `r Sys.time()`

This document was created using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com). Knitr allows the embedding of R code within markdown text documents allowing them to be updated and re-run. Things to note:

* Knitr will display warnings (but not errors) from R. The warnings may or may not be significant.
* Knitr is very clever but it does not always support pretty tables.

This code processes and anlayses the results of running https://github.com/dataknut/ping-log/blob/master/pinger.py. Pinger.py outputs a .csv file of the form:

    timestamp,host,milliseconds, error
    2016-04-28 10:53:56,www.google.co.uk,83.548, OK
    2016-04-28 10:53:57,router,121.820, OK
    2016-04-28 10:54:07,www.google.co.uk,71.019, OK
    2016-04-28 10:54:07,router,9.875, OK

Note that the R code will only process gzipped .csv files.

````{r, warning = FALSE}
# Housekeeping ----
# clear out all old objects etc to avoid confusion
rm(list = ls()) 


# set time
starttime <- Sys.time()

# set working directory
dpath <- "~/Dropbox/pinger/data/" # latest version of data with missing properly coded
rpath <- "~/Documents/pinger/results/" # where to put out output
tpath <- "~/Documents/pinger/tmp/" # where to put out output
gpath <- "~/Documents/pinger/results/" # where to put graphs

# load required packages ----
library(data.table) # fast data manipulation
library(foreign) # loading SPSS/STATA
library(ggplot2) # slick & easy graphs
library(gmodels) # for table proportions
library(knitr) # for kable

# set file name to load
# we may wish to load a number of them into different tables and then rbind
# but watch for timeing errors - e.g. RPi may be on UTC

#infile <- "octomac_annex_wifi_2016-04-28_10-53-56"
````

# Introduction
Purpose:

* To test connectivity to:
    + a home router and 
    + the wider internet 
* in order to attempt to work out where connectivity problems are occuring.

Data:

* Any number of .csv files produced by pinger.py

Code:

* this code: https://github.com/dataknut/ping-log/blob/master/pinger_process_results.Rmd

Warning:

* the code lines up the data files according to the datetime in them. If this is wrong (e.g. wrong timezone set or using UTC) then there will be mis-alignment.

# Load any pinger results files

You may see file read warnings below. They are usually caused by:
 
 * incorrect handling of ping response errors putting characters into the milliseconds field
 * unexpected file ending - e.g. when the pinger process quit without closing the file
 
All of these are re-coded as different types of 'error' before any data analysis.
 
```{r processRawData}
# may print warnings

# load just the one file
# pingerDT <- fread(paste0(dpath,infile,".csv"))

# Get file list and process
# This will only catch & load gzipped files (from now on)
# gzip them before downloading from the pinger - it's quicker & needs less storage space
# They load quicker too
setwd(dpath) # the glob function seems to fail if we give it the full path...
filelist <- list.files(pattern = glob2rx(paste0("*.csv.gz",
                                                sep = ""
                                                ),
                                         trim.head = FALSE, trim.tail = TRUE
                                         )
                       )

filesDT <- as.data.table(filelist) # makes for easy manipulation

# for each file in filelist we need to split on . and create the file source from the first word (before the .csv part)
# NB this assumes the filenames are meaningful!
filesDT$file <- sapply(strsplit(filesDT$filelist, "[.]"), "[[", 1) # why does R have such weird syntax for this?

# now get the unique file sources
uniqueSources <- unique(filesDT$filelist)

for(f in uniqueSources) {
  print(
    paste0("# Loading: ", f)
  )
  
  # Get the file using fread's ability to read from gzipped files
  # this may throw errors and warnings which we really should handle nicely but most of
  # them relate to poor ping error response parsing so text appears in the milliseconds column
  loadcmd <- paste0("gunzip -c ", f)
  temp_DT <- fread(loadcmd)
  
  names <- strsplit(f, "[.]" )[[1]] # split by . 
  source <- names[1] # first word in list = filename without suffix
  #print(
  #  paste0("# -> Setting source to: ", source)
  #)
  temp_DT$source <- source # set file name (without the .csv)

  
  # write out the table to the tmp folder ----
  # this is a bit of a kludge - but it allows the files to then be read in to one datatable later
  ofile <- paste0(tpath, source, "_DT.csv")
  #print(
  #  head(temp_DT)
  #)
  #print(
  #  summary(temp_DT)
  #)
  
  write.csv(temp_DT, ofile, row.names = FALSE)
  
  # create DT
  #dtname <- paste0(source, "_DT")
  #assign(dtname, temp_DT)
}

# remove temporary DT
temp_DT <- NULL
```

# Basic responses
Throughout the following NA usually means ping failed to return.

Files we processed:

````{r loadProcessedData}
setwd(tpath) # the glob function seems to fail if we give it the full path...
# Get file list and load
filelist <- list.files(pattern = glob2rx(paste0("*_DT.csv", 
                                                sep = ""
                                                ), 
                                         trim.head = FALSE, 
                                         trim.tail = TRUE
                                         )
                       )

print(filelist)

# now read them all in to one data table
allPinger_DT = as.data.table( #load as a data.table
  do.call(
    rbind, lapply(filelist, function(x) fread(x) # data.table fread function much quicker but prone to breaking if data formatting problems
    )
  )
)

# remove the temporary files
system("rm *_DT.csv")

#print("# -> Converting original date to R POSIXct")
allPinger_DT$r_datetime <- as.POSIXct(allPinger_DT$timestamp)
allPinger_DT[, timestamp := NULL] # remove 

#print("# -> adding date & hour variable")
allPinger_DT$r_date <- as.Date(allPinger_DT$r_datetime)
allPinger_DT$r_hour <- as.POSIXlt(allPinger_DT$r_datetime)$hour
````
We will now catch potential ping response errors as follows:

````{r, catchPingErrors, echo = TRUE}

# create uncaught error status ----
  # really nead to fix this in the pinger code
  # if word 'From' found in milliseconds column
  allPinger_DT$error <- ifelse(grepl("From",allPinger_DT$milliseconds),
                             "ping error: `From'",
                             allPinger_DT$error)
  # if record = OK but no milliseconds
  allPinger_DT$error <- ifelse(allPinger_DT$milliseconds == "" & allPinger_DT$error == "OK",
                             "ping error: no value returned",
                             allPinger_DT$error)
  # if record = OK & milliseconds is exactly 36, ping error (usually router not visible as away from home)
  allPinger_DT$error <- ifelse(allPinger_DT$milliseconds == "36" & allPinger_DT$error == "OK",
                             "ping error: '36'",
                             allPinger_DT$error)
  # if record = OK & milliseconds is exactly 92, ping error (usually router not visible as away from home)
  allPinger_DT$error <- ifelse(allPinger_DT$milliseconds == "92" & allPinger_DT$error == "OK",
                             "ping error: '92'",
                             allPinger_DT$error)

  # Force milliseconds to numeric (empty values will be set to NA)
  allPinger_DT[, milliseconds_numeric := as.numeric(milliseconds)]
  
  # Force any remaining NA milliseconds without error flags or 'OK' to have error flags
  allPinger_DT$error <- ifelse(is.na(allPinger_DT$milliseconds_numeric) &
                            allPinger_DT$error == "OK",
                             "ping error: unknown",
                             allPinger_DT$error)
  
  # Catch any remaining unset error codes
  allPinger_DT$error <- ifelse(is.na(allPinger_DT$error),
                             "error: unknown",
                             allPinger_DT$error)
  
  # Update milliseconds to numeric - this will force non-values to NA
  # Remember to exclude them later on!
  allPinger_DT[, milliseconds := milliseconds_numeric]
  allPinger_DT[, milliseconds_numeric := NULL]
  
```

We will now recode the sources as follows to make more meaningful labels for the charts etc:

````{r, recodeLabels, echo = TRUE}
# this assumes you know where they are and so what might cause any step changes in performance
# this is very hard to automate unless there is a naming convention for input files
allPinger_DT$label <- ifelse(grepl("hubpi",allPinger_DT$source),
                             "BThubNewHubpi",allPinger_DT$source) # copy in the source
allPinger_DT$label <- ifelse(grepl("pimine",allPinger_DT$source),
                             "BThubOldPimine",allPinger_DT$label)
allPinger_DT$label <- ifelse(grepl("hamishpi",allPinger_DT$source),
                             "hamishpi",allPinger_DT$label)
allPinger_DT$label <- ifelse(grepl("octomac",allPinger_DT$source),
                             "octomac",allPinger_DT$label)
allPinger_DT$label <- ifelse(grepl("ms_mbp",allPinger_DT$source),
                             "MrsA_mbp",allPinger_DT$label)
allPinger_DT$label <- ifelse(grepl("msmbp",allPinger_DT$source),
                             "MrsA_mbp",allPinger_DT$label)
````

How many rows (cases) & variables across all files?

````{r countsByHost, echo=FALSE}
dim(allPinger_DT)
kable(
  table(allPinger_DT$label, allPinger_DT$host, useNA = "always")
)
````

Distribution of milliseconds?

````{r millisecondsByLabel}
kable(
  allPinger_DT[,
               . (
                 Mean = round(mean(milliseconds, na.rm = TRUE),3),
                 N = length(milliseconds),
                 s.d = round(sd(milliseconds, na.rm = TRUE),3),
                 Last_reading = max(r_datetime)
               ),
               by = label
                 ]
)
````


# Ping error results

Did we get any errors?

````{r errorsByHost}
kable(
  table(allPinger_DT$error, allPinger_DT$label, useNA = "always")
)
````

We should have either error codes or 'OK'. If we have anything else then we have not caught all the errors properly earlier.

What time of day do we tend to get errors?

````{r errorsByHour, warning = FALSE}

# make a pretty graph of all errors by hour
# this will stack the counts
ggplot(allPinger_DT[allPinger_DT$error!="OK"], aes(x = r_hour)) + 
  geom_bar(aes(fill = error), position = "stack") +
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank()) +
  labs(x = "Hour",
       y = "Count"
  ) +
  facet_grid(label ~ .) +
  theme(strip.text.y = element_text(size = 8, colour = "red", angle = 0))
````

The bthub pinger seems to only see problems 15:00 - 16:00. As this Pi is 1 hour behind BST this equates to 16:00 - 17:00. The timing of errors for octomac (laptop) is probably not that relevant as this analysis includes data captured when it was outside the home.

The following chart counts errors by hour of each day to see if any particular days have been worse than others. Again ignore unkwon errors on octomac.

````{r errorsByDateHour, warning = FALSE}

# make a pretty graph of all errors by day & hour
allPinger_DT[, date_hour := as.POSIXct(paste0(r_date," ", r_hour, ":00:00"))]
# this will stack the counts
ggplot(allPinger_DT[allPinger_DT$error!="OK"], aes(x = date_hour)) + 
  geom_bar(aes(fill = error), position = "stack") +
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank()) +
  labs(x = "Date",
       y = "Count"
  ) +
  facet_grid(label ~ .) +
  scale_x_datetime(date_labels = "%a %d %b", date_breaks = "1 day") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(strip.text.y = element_text(size = 8, colour = "red", angle = 0))
````

So, most of the relevant errors happened on Friday 29th April. This was a) when the hail/thunderstorm hit at 16:00 and then b) when we subsequently called BT & were 'awarded' a new BTHub. This was installed on the 4th May. Who knows, maybe the old one got fried by the storm - or earlier? Either way, have things improved?

HamishPi, which sits on an internal ethernet segment fed by power line, clearly had problems at most times of day in late April - when we were re-configuring the power line ethernet.

# Ping response time results

The following graph shows all ping results for all sources between `r min(allPinger_DT$r_datetime)` and `r max(allPinger_DT$r_datetime)` for all data sources loaded. The dots represent the mean response time per hour and the error bars show +/- 2 s.d. as defined by the R [Hmisc package](http://www.inside-r.org/packages/cran/hmisc/docs/smean.sd). 

````{r millisecondsByDateTime, warning = FALSE, fig.cap = "All results for all days (mean & 2 * s.d.)"}
# make a pretty graph of all data by time

# create means per hour
ggplot(allPinger_DT, aes(x = date_hour, y = milliseconds)) + 
  # geom_point() + 
  #(stat = "mean", aes(color = host, col = "Response time (ms)")) +
  ggtitle("Mean response times (ms) per hour of each day") +
  aes(colour = host) +
  stat_summary(fun.data = "mean_sdl", size = 0.2) + # mean +/- 2 SD http://www.inside-r.org/packages/cran/hmisc/docs/smean.sd
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank()) +
  labs(x = "Date & time",
       y = "Milliseconds"
  ) +
  facet_grid(label ~ .) +
  scale_x_datetime(date_labels = "%a %d %b", date_breaks = "1 day") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(strip.text.y = element_text(size = 8, colour = "red", angle = 0))
````

Beware datetime issues on the Raspberry Pi - they may be one hour behind. This graph will get very big as more data is added and may become less useful as it does so.

Now we will repeat the analysis without aggregation for the last two days of data that are available and _only_ for the router Pi to be able to inspect the most recent results.

````{r millisecondsByDateTimeRecent, warning = FALSE, fig.cap = "HubPi results for the last two days (if we have any)"}
# repeat for yesterday & today
today <- Sys.Date()
yesterday <- today - 1

ggplot(allPinger_DT[(r_date == today | r_date == yesterday) & label %like% "BThub"], 
       aes(x = r_datetime, y = milliseconds)) +
  geom_point(aes(color = host, col = "Response time (ms)")) +
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank()) +
  labs(x = "Date & time",
       y = "Milliseconds"
  ) +
  facet_grid(label ~ .) +
  scale_x_datetime(date_labels = "%a %d %b %H:%M", date_breaks = "2 hours") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(strip.text.y = element_text(size = 8, colour = "red", angle = 0))
````

Guess what time the squids got home?

Fairly conclusively:

* there is rarely a problem with our broadband service - the RPi on the bthub generally shows low response times
* most of the problems are internal (no s**t Sherlock)

If we now consider the results by hour of the day:

```{r millisecondsByHour, warning = FALSE, fig.cap = "Mean milliseconds by hour"}
# make a pretty graph of all data by hour
resultsByHour_DT <- allPinger_DT[,
                                 .(
                                  mean_ms = round(mean(milliseconds, na.rm = TRUE),3)
                                 ),
                                 by = .(label,host,r_hour)
]  

ggplot(resultsByHour_DT, 
       aes(x = r_hour, y = mean_ms)) +
  geom_point(aes(color = host, col = "Response time (ms)")) +
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank()) +
  labs(x = "Date & time",
       y = "Mean milliseconds"
  ) +
  facet_grid(label ~ .) +
  theme(strip.text.y = element_text(size = 8, colour = "red", angle = 0))
````

Equally conclusively:

* the period 16:00 - 21:00 is, in general, the problem period. 
 
It is the _only_ problem period for the bthub (noting that for some reason, pimine_on_bthub's clock is 1 hour behind BST). Strangely enough this is also the period when the iPads-belonging-to-teenagers have internet access.

---------------------------------
Last run: `r Sys.time()`

Analysis & Knitr_ing completed in: `r Sys.time() - starttime` seconds using [knitr](https://cran.r-project.org/package=knitr) & [RStudio](http://www.rstudio.com)