---
title: "Analysis of pinger results"
author: "Ben Anderson (@dataknut)"
date: "28 April 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**About this document**

This document was created using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com). Knitr allows the embedding of R code within markdown text documents allowing them to be updated and re-run. Things to note:

* Knitr will display warnings (but not errors) from R. The warnings may or may not be significant.
* Knitr is very clever but it does not always support pretty tables.

This code processes and anlayses the results of running https://github.com/dataknut/ping-log/blob/master/pinger.py which is a .csv file with the form:

    timestamp,host,milliseconds, error
    2016-04-28 10:53:56,www.google.co.uk,83.548, OK
    2016-04-28 10:53:57,router,121.820, OK
    2016-04-28 10:54:07,www.google.co.uk,71.019, OK
    2016-04-28 10:54:07,router,9.875, OK

````{r, echo = FALSE, warning = FALSE}
# Housekeeping ----
# clear out all old objects etc to avoid confusion
rm(list = ls()) 


# set time
starttime <- Sys.time()

# set working directory
dpath <- "~/github/ping-log/results/" # latest version of data with missing properly coded

# load required packages ----
library(data.table) # fast data manipulation
library(foreign) # loading SPSS/STATA
library(ggplot2) # slick & easy graphs
library(gmodels) # for table proportions
library(knitr) # for kable

# set file name to load
# we may wish to load a number of them into different tables and then rbind
# but watch for timeing errors - e.g. RPi may be on UTC

infile <- "octomac_annex_wifi_2016-04-28_10-53-56"
````

# Introduction
Purpose:

* To test connectivity to:
 * a home router and 
 * the wider internet 
* in order to attempt to work out where connectivity problems are occuring.

Data:

* .csv files produced by pinger.py

Code:

* this code: https://github.com/dataknut/ping-log/blob/master/pinger_process_results.Rmd

```{r, echo=FALSE}
### Load data & do processing ---------------------------------

# load just the one file
# pingerDT <- fread(paste0(dpath,infile,".csv"))

# Get file list and process ----
setwd(dpath) # the glob function seems to fail if we give it the full path...
filelist <- list.files(pattern = glob2rx(paste0("*.csv",
                                                sep = ""
                                                ),
                                         trim.head = FALSE, trim.tail = TRUE
                                         )
                       )

filesDT <- as.data.table(filelist) # makes for easy manipulation

# for each file in filelist we need to split on . and create the file source
# NB this assumes the filenames are meaningful!
filesDT$file <- sapply(strsplit(filesDT$filelist, "[.]"), "[[", 1) # why does R have such weird syntax for this?

# now get the unique file sources
uniqueSources <- unique(filesDT$filelist)

# initiate processed list

processed <- ""

for(f in uniqueSources) {
  print(
    paste0("# Loading: ", f)
  )
  processed <- c(processed, f) # add ID to the list of those processed
  
  # Get the list of files for just this substation
  temp_DT <- fread(f)
  names <- strsplit(f, "[.]" )[[1]] # split by . 
  source <- names[1] # first word in list = filename without suffix
  #print(
  #  paste0("# -> Setting source to: ", source)
  #)
  temp_DT$source <- source # set file name (without the .csv)
  
  #print("# -> Converting original date to R POSIXct")
  temp_DT$r_datetime <- as.POSIXct(temp_DT$timestamp)

  # write out the table to the results folder
  # this is a bit of a kludge - but it allows the files to then be read in to one datatable later
  ofile <- paste0("temp/pinger_", source, "_DT.csv")
  print(
    head(temp_DT)
  )
  print(
    summary(temp_DT)
  )
  
  write.csv(temp_DT, ofile, row.names = FALSE)
  
  # create DT
  dtname <- paste0(source, "_DT")
  
  assign(dtname, temp_DT)
  

}

# remove temporary DT
temp_DT <- NULL
```

# Basic responses
Throughout the following NA usually means ping failed to return.

Files we processed:

````{r, echo=FALSE}
setwd(paste0(dpath,"temp/")) # the glob function seems to fail if we give it the full path...
# Get file list and load ----
filelist <- list.files(pattern = glob2rx(paste0("*_DT.csv", 
                                                sep = ""
                                                ), 
                                         trim.head = FALSE, 
                                         trim.tail = TRUE
                                         )
                       )

print(filelist)
allPinger_DT = as.data.table( #load as a data.table
  do.call(
    rbind, lapply(filelist, function(x) fread(x) # data.table fread function much quicker but prone to breaking if data formatting problems
    )
  )
)

#print("# -> Converting original date to R POSIXct")
allPinger_DT$r_datetime <- as.POSIXct(allPinger_DT$timestamp)
allPinger_DT$milliseconds <- as.numeric(allPinger_DT$milliseconds)

````

How many rows (cases) & variables across all files?

````{r, echo=FALSE}
dim(allPinger_DT)
kable(
  table(allPinger_DT$source, allPinger_DT$host, useNA = "always")
)
````

Did we get any errors?
````{r, echo=FALSE}
kable(
  table(allPinger_DT$source, allPinger_DT$error, useNA = "always")
)
````

# Results by host
Distribution of mean ping response times:
````{r, echo=FALSE}
kable(
  allPinger_DT[,
         .(
           N = length(r_datetime),
           Mean = mean(milliseconds, na.rm = TRUE), # remove any NAs (ping failed to respond)
           sd = sd(milliseconds, na.rm = TRUE)
           ),
         by = c("source", "host")
         ]
)
````

````{r, echo=FALSE}
# make a pretty graph
gtitle <- paste0("Ping Data: Key stats ", 
                 min(allPinger_DT$r_datetime), 
                 " to ",
                 max(allPinger_DT$r_datetime), 
                 " for ",
                 infile
                 )
ggplot(allPinger_DT, aes(x = r_datetime, y = milliseconds)) + 
  geom_point(aes(color = host, col = "Response time (ms)")) +
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank()) +
  labs(title = gtitle,
       x = "Time/date",
       y = "ms"
  ) +
  facet_grid(source ~ .)

ggsave(paste0(dpath, "all_data.png"), 
       width = 10, height = 10)
````


---------------------------------
Last run: `r Sys.time()`

Analysis completed in: `r Sys.time() - starttime` seconds using [knitr](https://cran.r-project.org/package=knitr) & [RStudio](http://www.rstudio.com)