---
title: "Analysis of pinger results"
author: "Ben Anderson (@dataknut)"
date: "28 April 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**About this document**

This document was created using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com). Knitr allows the embedding of R code within markdown text documents allowing them to be updated and re-run. Things to note:

* Knitr will display warnings (but not errors) from R. The warnings may or may not be significant.
* Knitr is very clever but it does not always support pretty tables.

This code processes and anlayses the results of running https://github.com/dataknut/ping-log/blob/master/pinger.py which is a .csv file with the form:

    timestamp,host,milliseconds
    2016-04-28 10:53:56,www.google.co.uk,83.548
    2016-04-28 10:53:57,router,121.820
    2016-04-28 10:54:07,www.google.co.uk,71.019
    2016-04-28 10:54:07,router,9.875

````{r, echo = FALSE, warning = FALSE}
# Housekeeping ----
# clear out all old objects etc to avoid confusion
rm(list = ls()) 


# set time
starttime <- Sys.time()

# set working directory
dpath <- "~/github/ping-log/results/" # latest version of data with missing properly coded

# load required packages ----
library(data.table) # fast data manipulation
library(foreign) # loading SPSS/STATA
library(ggplot2) # slick & easy graphs
library(gmodels) # for table proportions
library(knitr) # for kable

# set file name to load
# we may wish to load a number of them into different tables and then rbind
# but watch for timeing errors - e.g. RPi may be on UTC

infile <- "octomac_on_annex_wifi_2016-04-28_13-22-59"
````

# Introduction
Purpose:

* To test connectivity to:
 * a home router and 
 * the wider internet 
* in order to attempt to work out where connectivity problems are occuring.

Data:

* .csv files produced by pinger.py

Code:

* this code: https://github.com/dataknut/ping-log/blob/master/pinger_process_results.Rmd

```{r, echo=FALSE}
### Load data & do processing ---------------------------------

# load just the one file
pingerDT <- fread(paste0(dpath,infile,".csv"))

pingerDT$datetime_r <- as.POSIXct(pingerDT$timestamp)

# remove old datetime
pingerDT[,timestamp := NULL] # very quick
```

# Basic responses
Throughout the following NA usually means ping failed to return.

The file has `r length(pingerDT$datetime_r)` records starting at `r min(pingerDT$datetime_r)` and finishing at `r max(pingerDT$datetime_r)`.

How many rows (cases) & variables and how many records for each host?

````{r, echo=FALSE}
dim(pingerDT)

````

Did we get any errors?
````{r, echo=FALSE}
kable(
  table(pingerDT$error, pingerDT$host, useNA = "always")
)
````

Summary of variables (after converting datetime to R datetime (POSIX))
````{r, echo=FALSE}

summary(pingerDT)
````


# Results by host
Distribution of mean ping response times:
````{r, echo=FALSE}
pingerDT[,
         .(
           N = length(datetime_r),
           Mean = mean(milliseconds, na.rm = TRUE), # remove any NAs (ping failed to respond)
           sd = sd(milliseconds, na.rm = TRUE)
           ),
         by = host
         ]
````

````{r, echo=FALSE}
# make a pretty graph
gtitle <- paste0("Ping Data: Key stats ", 
                 min(pingerDT$datetime_r), 
                 " to ",
                 max(pingerDT$datetime_r), 
                 " for ",
                 infile
                 )
ggplot(pingerDT, aes(x = datetime_r, y = milliseconds)) + 
  geom_point(aes(color = host, col = "Response time (ms)")) +
  theme(legend.position = "bottom") +
  theme(legend.title = element_blank()) +
  labs(title = gtitle,
       x = "Time/date",
       y = "ms"
  )

ggsave(paste0(dpath,infile, ".png"), 
       width = 10, height = 10)
````

---------------------------------
Last run: `r Sys.time()`

Analysis completed in: `r Sys.time() - starttime` seconds using [knitr](https://cran.r-project.org/package=knitr) & [RStudio](http://www.rstudio.com)